Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes

### 2020-05-13 Ben_Manuscripts/hdphmm/Draft.tex
New explanations

  The ability to efficiently design novel membrane materials for selective separations
  would be greatly enhanced by easily interpretable methods for connecting microscopic
  fluctuations with macroscopic observables.
  \begin{itemize}
  	\item If one could apply molecular dynamics (MD) simulations to explain important
  	experimental metrics such as solute flux and selectivity in terms of detailed 
  	chemically dependent solute motion, then we can more easily bridge the disconnect
  	between theory and application.
  	\item In many systems, it's possible to achieve this by calculating diffusion 
  	constants.%\cite{}
  	\item Unfortunately, for systems which exhibit complex short timescale 
  	subdiffusive dynamics, one cannot calculate reliable diffusion 
  	constants since solute mean square displacements (MSDs) are non-linear 
    on the timescales accessible to molecular simulation.
  	\item Therefore, we are in need of a more flexible approach for modeling solute 
  	dynamics which makes less assumptions about the solute's average behavior.
  \end{itemize}

@@

  In this work we attempt to minimize human effort by applying the infinite hidden 
  Markov Model (IHMM)~\cite{beal_infinite_2002,calderon_inferring_2015}, a nonparameteric 
  Bayesian classification technique, in order to automatically detect and infer the 
  parameters of an unknown number of latent vector autoregressive (VAR) modes present 
  in solute center-of-mass time series trajectories. 
  \begin{itemize}
    %MRS: start with where it has been used before (to establish utility) a
    \item It's been used extensively for speaker diarization and has found some application
    in single particle trajectory analysis.~\cite{calderon_inferring_2015,fox_sticky_2011}
    %MRS: Then say it hasn't been used in this context (do we know of any, then cite, if none we have found, say so.)
    \item The IHMM has not been used much for MD trajectory analysis.

  \end{itemize}

@@

  Using the IHMM framework, we estimated the most likely sequence of hidden states in
  each solute center-of-mass trajectory while simultaneously inferring VAR(1)
  parameters for each state and the overall state transition probability matrix, $T$.
  We created a Python implementation of this process which we heavily adapted from
  the MATLAB code of Fox et al.~\cite{fox_bayesian_2010} Parameter estimation is iterative. 
  %MRS2: give location for the code here where you introduce it.
  %BJC3: I emailed Emily Fox and she just asked me to use this citation. I'm happy to add a 
  %link to where she stores the code. Would that be best in the form of a reference?
  %MRS3: I was thinking of citing YOUR code here.
  Therefore, we looked for convergence of the parameters in Equation~\ref{eqn:var}.
  We detected equilibration of the parameters using the module \texttt{pymbar.timeseries.detect\_equilibration}~\cite{chodera_simple_2016} 
  on the time series of parameter estimates. Plots illustrating parameter convergence are 
  available in Section~\ref{S-section:convergence} of the Supporting Information. We 
  refer the interested reader to much more extensive descriptions of the inference and 
  sampling procedures used to estimate the VAR(1) parameters and the state sequence. ~\cite{beal_infinite_2002,teh_hierarchical_2006,van_gael_beam_2008,fox_nonparametric_2009,fox_bayesian_2010}

  % BJC3: Cut this paragraph
%  We carefully applied the IHMM algorithm in a way which takes advantage of the
%  system's cylindrical symmetry. While there are a number of viable ways in which
%  one could choose to analyze these time series, we chose to use a multi-step 
%  procedure that we believe adequately distinguishes the diversity of dynamical 
%  behavior exhibited by the solutes. The final parameterization of the model is 
%  in cylindrical coordinates since each pore is a cylinder where we expect 
%  solutes to exhibit radially symmetric dynamics. We present a step-by-step 
%  demonstration of the procedure that follows in 
%  Section~\ref{S-section:ihmm_procedure} of the Supporting Information.

  We start by applying the IHMM algorithm to the 3D solute center-of-mass coordinate 
  trajectories transformed relative to the closest pore center. We tracked the solute's
  motion along the pore axis with the center-of-mass $z$ coordinate. Using the nearest
  pore center as the origin, we represented the radial distance of each solute's 
  center-of-mass from the pore center in 2 dimensions, $x$ and $y$. By working in 
  Cartesian coordinates, we avoid mathematical complexity introduced by change of coordinates into cylindrical 
  coordinates while estimating the state sequence.
  %BJC3: maybe could say something about multivariate Gaussian noise with VAR in cylindrical doesn't work with current code.
  %MRS3: tweaked above, probably OK.

@@

  % BJC3: This paragraph no longer necessary
%  We reparameterized the time series, preserving the state sequence, in terms
%  of the radial and axial coordinates ($r$, $z$) due to the system's radial 
%  symmetry. We converted the $x$ and $y$ center-of-mass coordinates to $r$. 
%  We fixed the state sequence from the 3D parameterization and applied the 
%  inference component of the IHMM procedure to the cylindrical trajectories 
%  in order to estimate the VAR(1) parameters in terms of $r$ and $z$. Since
%  we fixed the state sequence, and we are using conjugate priors, the VAR 
%  parameters converge very quickly (see Figure~\ref{S-fig:fixed_state_convergence}).
%  Therefore, we only ran the inference procedure for 100 iterations. 
%  % MRS: hmm. Here the new covariance parameters is redone in r - does there need to be any transformation applied? 
%  % MRS: though might be OK, since x^2 + y^2 = r^2 so maybe covariance are additive if x and y uncorrelated (which they should be)
%  % MRS2: As I mentioned before, the issue is if the likelihood for the covariance matrix is properly defined for radial diffusion.  We can check over the equations.  It likely wouldn't change the qualitative results much, but should ne checked.  Would not affect the clustering.
%  We defined the final parameters of each state as the mean of the parameters 
%  from each iteration recorded after the equilibration time point. We found the 
%  equilibration time point of each component of the $\mathbf{c}$ vector as well as 
%  $A$ and $\Sigma$ matrices, and then used the longest equilibration time of all 
%  dimensions as the equilibration time point.

  \subsubsection*{Predicting Mean Squared Displacement}\label{method:realizations}

  %MRS3: I think these belong after clustering.

  %BJC3: expanded
  \textit{Generating Stochastic IHMM Trajectory Realizations}: For each independent 
  MD trajectory, we generated stochastic trajectory realizations using the finalized 
  parameters of the associated IHMM. We drew state sequences with transition probabilities
  given by $T$. While in a given state, we simulated motion according to the VAR(1) 
  parameterization of that state. After each state transition, we set the unconditional
  mean of the new state to the particle's position immediately before the state 
  transition occurred.

  %BJC3: this is new, not sure of clarity
  \textit{Calculating MSD}: The MSD reported for each solute measured by our MD 
  simulations is a distribution created from the 24 independent MSD curves. We predicted 
  MSDs and their uncertainties with our stochastic models by bootstrapping the 
  distribution of stochastically generated MSDs. For each bootstrap trial, we generated
  24 independent trajectories using the 24 sets of IHMM parameters. We report the mean
  and standard error of the bootstrapped MSD distributions.

@@

  We reduced the parameter space used agglomerative clustering, a hierarchical
  clustering approach which uses a linkage criteria in order to successively merge
  similar clusters until a desired intracluster distance threshold or number of
  clusters is reached.~\cite{pedregosa_scikit-learn_2011} We used the Ward linkage 
  criteria, which works to minimize the sum of the squared differences within all
  clusters.~\cite{ward_hierarchical_1963} We elected to choose the number of clusters
  rather than the distance threshold. For our data, non-parametric methods such as 
  Bayesian Gaussian mixture models~\cite{pedregosa_scikit-learn_2011,gelman_bayesian_2013}
  tend to delocalize the clusters in parameter space (see Section~\ref{S-section:agglomerative}
  of the Supporting Information), which could cause the clustered model to combine 
  distinct behavior into a non-existent hybrid state.
  %MRS: not clear what the consequence of the above is?
  %BJC3: added a little to the end

  %BJC3: reworked the following and split into paragraphs
  We clustered based on the eigenvalues of $A$ and $\Sigma$, the distance from the 
  pore center and the self-transition probabilities, $T_{ii}$, of each state. $A$ 
  and $\Sigma$ capture the correlation and size of fluctuations. $T_{ii}$ is directly
  related to the average time spent in each state. The distance from the closest pore 
  center is likely meaningful since the membrane is radially inhomogeneous from each 
  pore, transitioning from hydrophilic pores to hydrophobic tails. However, the 
  channels are isotropic along the pore axis so we do not in means in the $z$ dimension.
  We normalized each clustering dimension by its variance and shifted the mean to zero.

  To take advantage of the system's cylindrical symmetry, we converted the parameters that
  are directly related to solute position into cylindrical coordinates before clustering. 
  %MRS3: tweaked: 
  %We converted the $x$ and $y$ eigenvalues of $\Sigma$ and $A$ as $\lambda_x$ + $\lambda_y$
  We replaced the $x$ and $y$ eigenvalues of $\Sigma$ and $A$ with
  $\lambda_x$ + $\lambda_y$ and $\lambda_x^2$ + $\lambda_y^2$
  respectively, which are invariant with rotation around the $z$ axis,
  in both cases reducing the dimensionality from 3 to 2.  We clustered
  on squared means in $r$, $\mu_r^2 = \mu_x^2 + \mu_y^2$, rather than
  $\mu_x$ and $\mu_y$ individually, with the square $\mu_r$ converting
  it into the same units as $A$ and $\Sigma$.

  To cluster on $T_{ii}$, we first cast it in terms of the expected dwell times. The expected
  value of the number of sequential self-transitions is simply $\frac{1}{1 - T_{ii}}$. This 
  relationship correctly implies that dwell times approach infinity as $T_{ii}$ approaches
  1. Therefore we found the most success with agglomerative clustering by linearizing this
  relationship and clustering on $-\log(1 - T_{ii})$ (see Section~\ref{S-section:selfT_cluster}
  of the Supporting Information for a graphical illustration).

%  One can choose alternative clustering
%  features, such as the eigenvalues of the $A$ and $\Sigma$ matrices, but we chose their 
%  diagonals because they can be easily translated to solute behavior. The diagonals of $A$
%  describe the correlation between sequential hops while the diagonals of $\Sigma$ describe
%  the size of the hops in each dimension. We assume that cross correlations in the parameters
%  are negligible for clustering.
%   
%  We also chose to cluster the diagonals of each matrix and the radial means
%  independently before combining the clusters to assign final state labels. If we divide
%  the total states into $m$ clusters based on the diagonal entries of $A$, $n$ more clusters 
%  based on the diagonal entries of $\Sigma$, and $k$ clusters based on $\mu_r$, then there
%  are a total of $mnk$ possible cluster combinations. This allows slightly more flexibility
%  in the total number of states compared to clustering on all five features at once, since 
%  there can be up to, but not necessarily exactly, $mnk$ states.

@@

  Choosing the number of clusters can be somewhat subjective so
  we attempted to add some structure to the selection process by following
  a set of qualitative and quantitative guidelines. We used the silhouette 
  test in order to score the quality of clustering as a function of the number 
  of clusters chosen.~\cite{kaufman_finding_2009} For our data, the silhouette 
  test generally favors the lowest number of clusters possible (see 
  Section~\ref{S-section:nclusters} of the Supporting Information). However, 
  choosing too few clusters tends to not distinguish between visually obvious 
  differences in dynamic behavior. This results in finalized parameter sets 
  that are averages of distinct behavior which presents further problems with 
  the predictive modeling that we discuss later on. We aimed to maintain the 
  highest silhouette score, and thus lowest number of clusters, possible while
  verifying that visually distinct states stayed separated.

@@

  We remapped and fixed the state sequence based on
  the cluster assignments and parameterized the clusters using the
  inference component of the IHMM algorithm.  First, we modified the
  ($x$, $y$, $z$) solute trajectories so that they had a mean of zero,
  leaving only the fluctuations. Next, we took advantage of the
  rotational invariance of the system by rotating the
  coordinates about the pore axis such that the mean vector of each
  separate segment was aligned with the $x$ axis. Next, we subtracted
  the means from each separate segment in the $x$ and $z$ dimensions,
  with the mean in $y$ being zeroed by the rotation.  We used the IHMM
  algorithm on this modified trajectory to infer the clustered state
  parameters by fixing the clustered state sequence.

  We obtained $\mathbf{c}$ vectors of the clustered states by averaging 
  each value of the rotated $\mathbf{c}$ vectors assigned to the same cluster. Note
  %MRS: tweaked
  that we only care about the $r=\sqrt{x^2+y^2}$ portion of $\mathbf{c}$ because solute 
  trajectories are not bound in the $z$ direction.

@@

  We quantified solute motion using the time averaged mean squared displacement (MSD)
  of the solute center of mass trajectories.~\cite{meroz_toolbox_2015} The time-averaged MSD measures all observed
  displacements over time lag $\tau$:
  \begin{equation}
  	\overline{z^2(\tau)} = \dfrac{1}{T - \tau}\int_{0}^{T - \tau} (z(t + \tau) - z(t))^2 dt
  \label{eqn:tamsd}
  \end{equation}
  where T is the length of the trajectory. We reported $1 \sigma$ confidence intervals 
  based on the results of 200 bootstrap trials. For each trail, we calculated the mean of 
  $n$ random trajectories chosen with replacement from the pool of trajectories, where 
  $n$ is the number of trajectories.

@@

  We applied the IHMM independently to all 24 trajectories of each of the four solutes studied.
  \begin{itemize}
    \item The number of states found for each trajectory varied between 5 and 30  %BJC2: need to see what these numbers actually are.
    \item In Figure~\ref{fig:rz_unclustered}, we show the state sequence determined by the
    IHMM for an example methanol trajectory.
    \item The model distinguished many of the states due to differences in their mean levels 
    while their fluctuations about those means are often similar. 
    \item In Section~\ref{section:mechanisms}, we use clustering to reduce the total state
    space for each solute.    
  \end{itemize}

  \begin{figure}
  \centering
  \includegraphics[width=\textwidth]{rz_unclustered_MET.pdf}
  \caption{The IHMM found 24 distinct VAR(1) states in this representative methanol trajectory.}\label{fig:rz_unclustered}
  \end{figure}

  \subsection{Reproducing MD Trajectories and MSDs with the IHMM}

  % MRS: maybe makes sense to talk a little about what features/parameters of the 
  % models the MSD is sensitive to?

  Before analyzing solute behavior characteristic to the states identified by the
  IHMM, it is important to verify whether the state dynamics are consistent with MD
  both qualitatively and quantitatively.
  \begin{itemize}
    \item Therefore, we generated stochastic realizations based on the parameters 
    of each trajectory as described in Section~\ref{method:realizations}.
  \end{itemize}

  As shown in Figure~\ref{fig:qualitative_unclustered}, realizations of our model 
  can produce trajectories that show qualitatively similar hopping and trapping 
  behavior to the MD trajectories.
  \begin{itemize}
    \item It is important that our model reproduces this behavior so that we can
    have confidence in its usefulness for quantitative predictions.
  \end{itemize}

  \begin{figure}
  \centering
  \includegraphics[width=\textwidth]{qualitative_unclustered_MET2.pdf}
  \caption{We can generate trajectories which bear qualitative resemblance to
  the MD trajectories. Solute trajectories generated by the IHMM (blue) show
  the same hopping and trapping behavior exhibited by MD (black).
  }\label{fig:qualitative_unclustered}
  \end{figure}

  \noindent From a quantitative standpoint, the model has two shortcomings.
  \begin{itemize}
  	\item First, in many cases, the MSD predicted by realizations of the IHMM 
  	under-predicts that of MD (see Figure~\ref{fig:unclustered_msds}).
  	\item Second, the shapes of the MSD curves at short time lags are very different.
  \end{itemize}

  \begin{figure}
  \centering
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_MET.pdf}
  \caption{methanol}\label{fig:unclustered_msd_MET}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_GCL.pdf}
  \caption{ethylene glycol}\label{fig:unclustered_msd_GCL}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_URE.pdf}
  \caption{urea}\label{fig:unclustered_msd_URE}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_ACH.pdf}
  \caption{acetic acid}\label{fig:unclustered_msd_ACH}
  \end{subfigure}
  \caption{The MD MSDs (black) of methanol and urea are under-predicted by 
  the IHMM (orange), while the predictions match within uncertainty for ethylene
  glycol and urea.}\label{fig:unclustered_msds}
  %MRS2: to save space, could consider the graphs in 1x4 across the top, with a single legend (since all have the same), and slightly thicker lines  %There are only 2 lines per plot, so it's not too noisy to be that size
  %BJC3: good point
  \end{figure}

  %MRS2: I'd make this a separate subsection ``Deviations of the IMMM from molecular motion'' or something like that.
  % BJC3: would these subsubsection headers look better as just italicized and on 
  % the same line as where the next paragraph starts?
  %MRS3: yes, that's what I'm thinking of.
  \subsubsection*{Deviations of the IHMM from Molecular Motion}

  \noindent The IHMM under-predicts MD MSD due to its tendency to favor fewer states.%\cite{}
  \begin{itemize}
    \item Disjoint segments with similar VAR parameters, but different lengths of 
    entrapment can be parameterized as the same state.
    \item In Figure~\ref{fig:dwell_distributions}, we compare the overall dwell 
    time distributions generated directly from MD simulations to that generated from 
    realizations of the IHMM. 
    \item Some of the probability density of short trapping times in the MD simulations
    gets shifted to medium length trapping times in realizations of the IHMM.
    \item In Figure~\ref{fig:dwell_distributions_modT} and~\ref{fig:msd_modT}, we show 
    that by modifying the transition matrix to encourage transitions towards states with
    shorter dwell times, we can more closely reproduce the dwell time distribution and 
    MSD of MD.
    \item This is because most solute motion is a consequence of transitions between 
    trapped states.
  \end{itemize}

  \begin{figure}
  \centering
  \begin{subfigure}{0.32\textwidth}
  \includegraphics[width=\textwidth]{dwell_distributions.pdf}
  \caption{}\label{fig:dwell_distributions}
  \end{subfigure}
  \begin{subfigure}{0.32\textwidth}
  \includegraphics[width=\textwidth]{dwell_distributions_modT.pdf}
  \caption{}\label{fig:dwell_distributions_modT}
  \end{subfigure}
  \begin{subfigure}{0.32\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_MET_modT.pdf}
  \caption{}\label{fig:msd_modT}
  \end{subfigure}
  \caption{(a) The density of short dwell times compiled from realizations of
  the IHMM is below that measured directly from the MD trajectories. If we
  modify the transition matrix in order to favor transitions towards states with
  short dwell times, we can more closely match the (b) dwell time distributions
  and (c) MSDs.}\label{fig:slow_T}
  \end{figure}

>>>>

  The ability to efficiently design novel membrane materials for selective separations
  would be greatly enhanced by easily interpretable methods for connecting microscopic
  fluctuations with macroscopic observables. If one could apply molecular dynamics 
  (MD) simulations to explain important experimental metrics such as solute flux and 
  selectivity in terms of detailed chemically dependent solute motion, then we can more
  easily bridge the disconnect between theory and application. In many systems, it's 
  possible to achieve this by calculating diffusion constants. Unfortunately, for 
  systems which exhibit complex short timescale subdiffusive dynamics, one cannot 
  calculate reliable diffusion constants since solute mean square displacements (MSDs)
  are non-linear on the timescales accessible to molecular simulation. Therefore, we
  are in need of a more flexible approach for modeling solute dynamics which makes 
  less assumptions about the solute's average behavior.

@@

  In this work we attempt to minimize human effort by applying the infinite hidden 
  Markov Model (IHMM)~\cite{beal_infinite_2002,calderon_inferring_2015}, a nonparameteric 
  Bayesian classification technique, in order to automatically detect and infer the 
  parameters of an unknown number of latent vector autoregressive (VAR) modes present 
  in solute center-of-mass time series trajectories. 
  \begin{itemize}
    %MRS: start with where it has been used before (to establish utility) a
    \item It's been used extensively for speaker diarization and has found some application
    in single particle trajectory analysis.~\cite{calderon_inferring_2015,fox_sticky_2011}
    %MRS: Then say it hasn't been used in this context (do we know of any, then cite, if none we have found, say so.)
    \item The IHMM has not been used much for MD trajectory analysis.
  \end{itemize}

@@

  Using the IHMM framework, we estimated the most likely sequence of hidden states in
  each solute center-of-mass trajectory while simultaneously inferring VAR(1)
  parameters for each state and the overall state transition probability matrix, $T$.
  We created a Python implementation of this process 
  (\texttt{https://github.com/bencoscia/hdphmm}) which we heavily adapted from
  the MATLAB code of Fox et al.~\cite{fox_bayesian_2010} Parameter estimation is iterative. 
  %MRS2: give location for the code here where you introduce it.
  %BJC3: I emailed Emily Fox and she just asked me to use this citation. I'm happy to add a 
  %link to where she stores the code. Would that be best in the form of a reference?
  %MRS3: I was thinking of citing YOUR code here.
  %BJC4: Ah, OK. I plan to fork it into the shirtgroup repo after we submit. Can change link after reviews.
  %BJC4: Okay to put link? Or should I cite it as a reference?
  Therefore, we looked for convergence of the parameters in Equation~\ref{eqn:var}.
  We detected equilibration of the parameters using the module \texttt{pymbar.timeseries.detect\_equilibration}~\cite{chodera_simple_2016} 
  on the time series of parameter estimates. Plots illustrating parameter convergence are 
  available in Section~\ref{S-section:convergence} of the Supporting Information. We 
  refer the interested reader to much more extensive descriptions of the inference and 
  sampling procedures used to estimate the VAR(1) parameters and the state sequence. ~\cite{beal_infinite_2002,teh_hierarchical_2006,van_gael_beam_2008,fox_nonparametric_2009,fox_bayesian_2010}

  We start by applying the IHMM algorithm to the 3D solute center-of-mass coordinate 
  trajectories transformed relative to the closest pore center. We tracked the solute's
  motion along the pore axis with the center-of-mass $z$ coordinate. Using the nearest
  pore center as the origin, we represented the radial distance of each solute's 
  center-of-mass from the pore center in 2 dimensions, $x$ and $y$. By working in 
  Cartesian coordinates, we avoid mathematical complexity introduced by change of 
  coordinates into cylindrical coordinates while estimating the state sequence.

@@

%  It is important not to view the IHMM as a black box. One should verify that the
%  model produces qualitative and quantitative results that are accurate on a
%  trajectory-by-trajectory basis.
%  \begin{itemize}
%    \item If the state sequence is not properly partitioned it can lead to 
%    poor estimates of the transition matrix.
%    \item In the supporting information, we demonstrate this and provide techniques
%    for ensuring good state partitioning.
%  \end{itemize}

@@

  We reduced the parameter space used agglomerative clustering, a hierarchical
  clustering approach which uses a linkage criteria in order to successively merge
  similar clusters until a desired intracluster distance threshold or number of
  clusters is reached.~\cite{pedregosa_scikit-learn_2011} We used the Ward linkage 
  criteria, which works to minimize the sum of the squared differences within all
  clusters.~\cite{ward_hierarchical_1963} We elected to choose the number of clusters
  rather than the distance threshold. For our data, non-parametric methods such as 
  Bayesian Gaussian mixture models~\cite{pedregosa_scikit-learn_2011,gelman_bayesian_2013}
  tend to delocalize the clusters in parameter space (see Section~\ref{S-section:agglomerative}
  of the Supporting Information), which could cause the clustered model to combine 
  distinct behavior into a non-physical hybrid state.

  We clustered based on the eigenvalues of $A$ and $\Sigma$, the distance from the 
  pore center and the self-transition probabilities, $T_{ii}$, of each state. $A$ 
  and $\Sigma$ capture the correlation and size of fluctuations. $T_{ii}$ is directly
  related to the average time spent in each state. The distance from the closest pore 
  center is likely meaningful since the membrane is radially inhomogeneous from each 
  pore, transitioning from hydrophilic pores to hydrophobic tails. However, the 
  channels are isotropic along the pore axis so we do not in means in the $z$ dimension.
  We normalized each clustering dimension by its variance and shifted the mean to zero.

  To take advantage of the system's cylindrical symmetry, we converted the parameters that
  are directly related to solute position into cylindrical coordinates before clustering. 
  We replaced the $x$ and $y$ eigenvalues of $\Sigma$ and $A$ with $\lambda_x$ + 
  $\lambda_y$ and $\lambda_x^2$ + $\lambda_y^2$ respectively, which are invariant with 
  rotation around the $z$ axis, in both cases reducing the dimensionality from 3 to 2.  
  We clustered on squared means in $r$, $\mu_r^2 = \mu_x^2 + \mu_y^2$, rather than
  $\mu_x$ and $\mu_y$ individually, with the square $\mu_r$ converting it into the 
  same units as $A$ and $\Sigma$.

  To cluster on $T_{ii}$, we first cast it in terms of the expected dwell times. The expected
  value of the number of sequential self-transitions is simply $\frac{1}{1 - T_{ii}}$. This 
  relationship correctly implies that dwell times approach infinity as $T_{ii}$ approaches
  1. Therefore we found the most success with agglomerative clustering by linearizing this
  relationship and clustering on $-\log(1 - T_{ii})$.
  % (see Section~\ref{S-section:selfT_cluster} of the Supporting Information for a graphical illustration).

  %BJC4: Added
  There are alternative ways to cluster the data dependent on one's goals. Our goal is
  to derive the simplest set of parameter which adequately describe solute motion. 
  However, one may want a more detailed understand of specific dimensions of the 
  clustering dataset. In this case one can cluster on one or more dimensions independently
  and then enumerate the possible clusters. This would ensure very little overlap of
  clusters but may result in a large number of them. Once one has achieved a better 
  understanding of solute-membrane interactions, perhaps by using the iHMM, it may be
  beneficial to incorporate other physically motivated variables into the clustering
  procedure. For example, one can gain a thorough understanding of the dynamics of
  different kinds of trapping behavior related to hydrogen bonding and sodium ion association.

@@

  Choosing the number of clusters can be somewhat subjective so we attempted to add some
  structure to the selection process by following a set of qualitative and quantitative 
  guidelines. We give an in depth justification of our clustering choices in 
  Section~\ref{S-section:nclusters} of the Supporting Information. We used the silhouette
  test in order to score the quality of clustering as a function of the number of clusters
  chosen.~\cite{kaufman_finding_2009} For our data, the silhouette test generally favors 
  the lowest number of clusters possible. However, choosing too few clusters tends to not
  distinguish between visually obvious differences in dynamic behavior. This can result 
  in finalized parameter sets that are averages of distinct behavior which presents 
  further problems with the predictive modeling that we discuss later on.

  %BJC4: added
  We aimed to maintain the highest silhouette score, and thus lowest number of clusters,
  possible while preserving the dynamics of each trajectory.
  %verifying that visually distinct states stayed separated. 
  We verified preservation of the dynamics by comparing the MSDs of stochastic trajectory
  realizations based on clustered IHMM parameter sets to the pooled MSDs of stochastic 
  realizations of IHMMs fit to individual trajectories (see Section~\ref{method:realizations}).
  We chose the lowest number of clusters that reasonably reproduced the pooled MSDs at 
  short and long time lags (See Figure~\ref{S-fig:MET_nclusters} of the Supporting 
  Information).

@@

  We remapped and fixed the state sequence based on the cluster assignments and 
  parameterized the clusters using the inference component of the IHMM algorithm.
  First, we modified the ($x$, $y$, $z$) solute trajectories so that they had a 
  mean of zero, leaving only the fluctuations. Next, we took advantage of the
  rotational invariance of the system by rotating the coordinates about the pore 
  axis such that the mean vector of each separate segment was aligned with the 
  $x$ axis. Next, we subtracted the means from each separate segment in the $x$ 
  and $z$ dimensions, with the mean in $y$ being zeroed by the rotation. We used
  the inference component of the IHMM algorithm on this modified trajectory to 
  infer the clustered state parameters by fixing the clustered state sequence.

  We obtained $\mathbf{c}$ vectors of the clustered states by averaging each 
  value of the rotated $\mathbf{c}$ vectors assigned to the same cluster. Note
  that we only care about the $r=\sqrt{x^2+y^2}$ portion of $\mathbf{c}$ because
  solute trajectories are not bound in the $z$ direction.

  \subsection{Predicting Mean Squared Displacement}\label{method:realizations}

  \textit{Generating Stochastic IHMM Trajectory Realizations}: For each independent 
  MD trajectory, we generated stochastic trajectory realizations using the finalized 
  parameters of the associated IHMM. We drew state sequences with transition probabilities
  given by $T$. While in a given state, we simulated motion according to the VAR(1) 
  parameterization of that state. After each state transition, we set the unconditional
  mean of the new state to the particle's position immediately before the state 
  transition occurred.

  \textit{Calculating MSD}: We quantified solute motion using the time averaged mean squared displacement (MSD)
  of the solute center of mass trajectories.~\cite{meroz_toolbox_2015} The time-averaged MSD measures all observed
  displacements over time lag $\tau$:
  \begin{equation}
  	\overline{z^2(\tau)} = \dfrac{1}{T - \tau}\int_{0}^{T - \tau} (z(t + \tau) - z(t))^2 dt
  \label{eqn:tamsd}
  \end{equation}
  where T is the length of the trajectory. We reported $1 \sigma$ confidence intervals 
  based on the results of 200 bootstrap trials. For each trail, we calculated the mean of 
  $n$ random trajectories chosen with replacement from the pool of trajectories, where 
  $n$ is the number of trajectories.

  The MSD reported for each solute measured by our MD 
  simulations is a distribution created from the 24 independent MSD curves. We predicted 
  MSDs and their uncertainties with our stochastic models by bootstrapping the 
  distribution of stochastically generated MSDs. For each bootstrap trial, we generated
  24 independent trajectories using the 24 sets of IHMM parameters. We report the mean
  and standard error of the bootstrapped MSD distributions.

@@

@@

  We applied the IHMM independently to all 24 trajectories of each of the four solutes studied.
  The number of states found for each trajectory varied between 5 and 30.  %BJC2: MET: (5, 28)
  In Figure~\ref{fig:rz_unclustered}, we show the state sequence determined by the IHMM for
  an example methanol trajectory. The model distinguishes many of the states due to 
  differences in their mean levels while their fluctuations about those means are often 
  similar. In Section~\ref{section:mechanisms}, we use clustering to reduce the total state
  space for each solute in order to more easily interpret the state parameters.

  %BJC4: It might be good to add a panel to this figure which gives a radial view of the pore
  \begin{figure}
  \centering
  \includegraphics[width=\textwidth]{rz_unclustered_MET.pdf}
  \caption{The IHMM found 24 distinct VAR(1) states in this representative methanol trajectory.
  We performed the parameterization in 3D cartesian coordinates, but for this figure, we 
  converted to cylindrical coordinates, with $r$ the distance from the pore center, to 
  provide a clearer picture of solute motion.}\label{fig:rz_unclustered}
  \end{figure}

  \subsection{Reproducing MD Trajectories and MSDs with the IHMM}\label{section:unclustered_MSD_prediction}

  Before analyzing solute behavior characteristic to the states identified by the
  IHMM, it is important to verify whether the state dynamics are consistent with MD
  both qualitatively and quantitatively. Therefore, we generated stochastic 
  realizations based on the parameters of each trajectory as described in 
  Section~\ref{method:realizations}.

  As shown in Figure~\ref{fig:qualitative_unclustered}, realizations of our model 
  can produce trajectories that show qualitatively similar hopping and trapping 
  behavior to the MD trajectories. It is important that our model reproduces this 
  behavior so that we can have confidence in its usefulness for quantitative 
  predictions.

  % BJC4: This figure needs to be redone with new trajectory generation procedure
  \begin{figure}
  \centering
  \includegraphics[width=\textwidth]{qualitative_unclustered_MET2.pdf}
  \caption{We can generate trajectories which bear qualitative resemblance to
  the MD trajectories. Solute trajectories generated by the IHMM (blue) show
  the same hopping and trapping behavior exhibited by MD (black).
  }\label{fig:qualitative_unclustered}
  \end{figure}

  \textit{Deviations of the IHMM from Molecular Motion}: From a quantitative
  standpoint, using the IHMM as a predictive tool has two 
  shortcomings. First, in many cases, the MSD predicted by realizations of the 
  IHMM under-predicts that of MD (see Figure~\ref{fig:unclustered_msds}).
  Second, the shapes of the MSD curves at short time lags are very different.

  \begin{figure}
  \centering
  \begin{subfigure}{0.24\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_MET.pdf}
  \caption{methanol}\label{fig:unclustered_msd_MET}
  \end{subfigure}
  \begin{subfigure}{0.24\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_GCL.pdf}
  \caption{ethylene glycol}\label{fig:unclustered_msd_GCL}
  \end{subfigure}
  \begin{subfigure}{0.24\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_URE.pdf}
  \caption{urea}\label{fig:unclustered_msd_URE}
  \end{subfigure}
  \begin{subfigure}{0.24\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_ACH.pdf}
  \caption{acetic acid}\label{fig:unclustered_msd_ACH}
  \end{subfigure}
  \caption{The MD MSDs (black) of methanol and urea are under-predicted by 
  the IHMM (orange), while the predictions match within uncertainty for ethylene
  glycol and acetic acid}\label{fig:unclustered_msds}
  \end{figure}

  The IHMM under-predicts the MD MSD due to its tendency to favor fewer states. 
  Disjoint segments with similar VAR parameters, but different lengths of entrapment
  can be parameterized as the same state. In Figure~\ref{fig:dwell_distributions}, 
  we compare the overall dwell time distributions generated directly from MD simulations to that 
  generated from realizations of the IHMM. We do this by measuring the time spent in 
  each independent same-state segment. Some of the probability density of short 
  trapping times in the MD simulations gets shifted to medium length trapping times
  in realizations of the IHMM. In Figure~\ref{fig:dwell_distributions_modT} 
  and~\ref{fig:msd_modT}, we show that by modifying the transition matrix to 
  encourage transitions towards states with shorter dwell times, we can more closely
  reproduce the dwell time distribution and MSD of MD. This is because most solute
  motion is a consequence of transitions between trapped states.

  % BJC4: A lower priority supporting information section I've had in mind is one
  % with tips for fitting and validating the IHMM on individual trajectories.  
%  In many cases, especially evident where MSD predictions over-estimate MD, there
%  is some error in the state sequence segmentation. In Section~\ref{S-section:fitting_tips}
%  of the Supporting Information, we give some tips based on our experience of how to 
%  initialize IHMM fits in order to reduce error.

  \begin{figure}
  \centering
  \begin{subfigure}{0.32\textwidth}
  \includegraphics[width=\textwidth]{dwell_distributions.pdf}
  \caption{}\label{fig:dwell_distributions}
  \end{subfigure}
  \begin{subfigure}{0.32\textwidth}
  \includegraphics[width=\textwidth]{dwell_distributions_modT.pdf}
  \caption{}\label{fig:dwell_distributions_modT}
  \end{subfigure}
  \begin{subfigure}{0.32\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_MET_modT.pdf}
  \caption{}\label{fig:msd_modT}
  \end{subfigure}
  \caption{(a) The density of short dwell times compiled from realizations of
  the IHMM fit to methanol trajectories is below that measured directly from 
  the MD trajectories. If we modify the transition matrix in order to favor 
  transitions towards states with short dwell times, we can more closely match
  the (b) dwell time distributions and (c) MSDs.}\label{fig:slow_T}
  \end{figure}
